---
title: "Bird Point Counts"
description: |
  Traditional morning point counts
output:
    distill::distill_article:
      toc: true
      toc_depth: 3
---
NOTE: I severely reduced the sample dataset because I saw a suspicious amount of Github repo clones and don't want the data everywhere. I will remake figures locally using a larger subset of data and upload them as files.


Each sampling point had two bird point counts, on consecutive days. The point counts were conducted starting 15 minutes before sunrise, and 


There are two csv files for this analysis, one with deployment metadata and one with bird point count data: 

```{r echo=FALSE}
#xfun::embed_file('docs/data/ETH dry season deployments AZ001-053.csv')
```  

```{r echo=FALSE}
#xfun::embed_file('docs/data/ETH dry season birds AZ001-053.csv')
```  



Here is how I set up the environment and did some initial data cleanup:

<details markdown="1">
<summary><strong>Click here</strong> for code </summary>

```{r setup, message=FALSE, results = 'hide'}
setwd("C:/Users/Hubert/Desktop/Biodiversity Monitoring/Biodiversity data analysis/Biodiversity-analysis")

#Load libraries
library(ggplot2)
library(vegan)
library(dplyr)
library(viridis) 
library(tidyr)

# Load the datasets
metadata <- read.csv('media/data/ETH dry season deployments AZ001-053.csv')
data <- read.csv('media/data/ETH dry season birds AZ001-053.csv')

#data cleanup

#remove rows/columns not in analysis
metadata <- metadata[c(1:100),]                 ### which points are we analyzing?
data <- data[c(1:20)]                           ### extra blank columns
data<-data[which(complete.cases(data$date)),]   ### extra blank rows

## just in case confirmations:

#making sure there are no misspelled names - scroll through and see if there are two in a row only a letter off
#genus_species <- unique(data$bird)
#species_list <-sort(genus_species)
#species_list <- as.data.frame(species_list)

#make sure there are two counts per each point - scroll through and see if there are two for each point
#counts <- unique(data$bird_count)
#counts <-sort(counts)
#counts <- as.data.frame(counts)

### 12.6.25 - we are missing PON26-50; 68-84, so only have 58 of the 100 points

#remove blank rows (if the number of rows changes in data, then there are blanks, look into why)
#data[data == ""] <- NA  #replace blank with NA
#data <- data[complete.cases(data$name), ]

```

</details>

Each point count consisted of 10 minutes of observation, followed by 5 minutes of attracting birds with a playback recording. 

For now, I will only consider the 10 minute point count, so I split the data here:

<details markdown="1">
<summary><strong>Click here</strong> for code </summary>

```{r, message=FALSE, results = 'hide'}
#split bird data into playback and non-playback
bird_playback_data <- subset(data, playback != "no",)
bird_point_data <- subset(data, playback == "no",)
```
</details>

Now, here is an important inflection point. The 10 minute point counts were conducted in 2-minute sub-sampling blocks. For example, if a bird was observed in the first two minutes, and in the second two minutes, but not in minutes 4-10, then they get a '1' in data$X0 and data$X2, but not in data$X4,X6, and X8. This data entry system was used to allow a mark-recapture style analysis with higher precision in abundance values. This is based on the approach from Alldredge et al (2007) https://doi.org/10.1093/auk/124.2.653. Tom BradferLawrence, who recommended that I do this kind of sampling, says "Abundance modelling which accounts for differences in detectability is done using a Bayesian hierarchical framework", and he recommends looking into the spAbundance R package (Doser et al).

However, all that is complicated, so I am not going to do that at the moment, and just use number of rows as abundance (same species, different individuals get their own rows).
